---
title: "Module 2: Overfishing"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

```{r}
#| warning: false
library(tidyverse)
```


### Use of GitHub

Link to your forked GH repository: https://github.com/tbaugh323/Module2-overfishing.git

### Use of Quarto

Link to your .qmd file: https://github.com/tbaugh323/Module2-overfishing/blob/main/module2.qmd

## Overexploitation of global fisheries

We will examine the state of global fisheries and seek to reproduce one of the most widely cited examples of species collapse in order to examine the evidence behind an influential paper on global fisheries, [Worm et al. 2006](https://doi.org/10.1126/science.1132294). However, rather than using the 1950-2003 dataset available to Worm and colleagues in 2006, we will be drawing from more recent stock assessment data to see how the trends have fared, specifically in Atlantic cod.

We also explore how to understand and manipulate tabular data using relational database concepts. Instead of working with a single independent dataframe, as in Module 1, we will be working with a large relational database consisting of tables of different dimensions, but are related through one or more IDs.

## The Database

We will use data from the [RAM Legacy Stock Assessment Database](https://www.ramlegacy.org/database/). Note that the database files are available through Zenodo; please download and unzip the file, moving the .RData file only into your own `data/` folder within this project.

Use the code below to load up the tables in this database.

```{r}
load("data/DBdata[asmt][v4.66].RData")
```

The above function loads all 60 tables of this database, which is a lot! So step one of the data science process is exploring the structure of the database and figuring out how the tables are related.

A selected set of documentation have been provided in `documentation/`. Open up the Database Quickstart Guide for reference as you work through the problems below.

### Q1 (1 point)

One of the main metadata tables is 'stock':

-   How many stocks are there and how many species do they represent? Either common or scientific names okay, although they give different answers. 

-   What are the top three most common species? (Okay to use common names; these are mostly standardized for fish)

-   In your own words, describe the unit of observation in the 'stock' table and some associated variables that describe it.

```{r}
num_stocks <- length(unique(stock$stockid))

num_species <- length(unique(stock$commonname))

stock |> 
  group_by(commonname) |> 
  summarize(count = n()) |> 
  arrange(desc(count))

```

There are 1514 stocks that represent 398 different species. The three most common species are pink salmon, sockeye salmon, and chum salmon. Each stock observation represents a stock with a unique stockid and information on what it represents, such as species and location.

### Q2 (1 point)

The next main metadata table is 'assessment':

-   What is the unit of observation in the 'assessment' table? How is this related to the unit of observation in the 'stock' table?

-   It seems like 'stockid' and 'assessyear' might uniquely identify each assessment. Is this the case? If not, why not?

-   Which variable(s) might you use to join 'stock' and 'assessment'?

```{r}
assessment |> 
  group_by(stockid, assessyear) |> 
  summarize(count = n()) |> 
  summary()

View(assessment |> 
  group_by(stockid, assessyear) |> 
  summarize(count = n()))
```

The unit of observation is the assessment ID, which describes the assessment method. The stock table unit of observation was each individual stock ID, which the assessment table also has. So, each stock observation is further described in the assessment table with the assessment method. stockid and assessyear do not uniquely identify each assessment, as there are several stock IDs which were assessed twice in one year. We'd join stock and assessment together with the stockid and stocklong variables.

### Q3 (1 point)

In section B of the Database Quick Guide, additional metadata tables are listed in section B. With your data exploration skills, complete the following table:

| B table      | Associated table | join column from B | join column from other |
|------------------|------------------|------------------|------------------|
| area         |                  |                    |                        |
| assessmethod |                  |                    |                        |
| assessor     |                  |                    |                        |
| management   |                  |                    |                        |
| taxonomy     |                  |                    |                        |

\*Note that assessmethod is not a data.frame, which was perhaps an oversight; consider turning it into one for the ease of using tidyverse functions.

### Q4 (1 point)

There are two main data tables described in section C, but their key attributes are difficult to understand. Join the two main data tables with their respective metrics tables (section D):

-   Briefly describe the contents of each joined table.

-   What is accomplished by joining these tables?

```{r}
bio_all <- biometrics |> 
  left_join(bioparams, by = join_by(biounique == bioid))

time_all <- timeseries |> 
  left_join(tsmetrics, by = join_by(tsid == tsunique))
```

The biometrics and bioparams tables have a bunch of biological traits for each stock for each time it was assessed. The timeseries and timemetrics tables have a lot of time-related parameters for each stock id each time it was assessed. By joining these tables, we have all the information about these parameters and the stocks they are associated with in one place.

## Overexploitation of Atlantic cod

This database is clearly complex and contains a lot of information. We've grappled with its high dimensionality by exploring and joining tables above. Next, we will take a "bottom-up" approach and evaluate a single species, Atlantic cod. Read this [book chapter](https://pressbooks.pub/extinctionstories/chapter/atlantic-cod/) on Atlantic cod, which serves as a reference for the following questions.

### Q5 (1 point)

The collapse of the Atlantic cod fisheries is well documented. Let's see if we can visualize that decline with this database. 

-   How many unique fish stocks comprise "Atlantic cod" and what regions do they come from?


Create a vector of the 'stockid' associated with "Atlantic cod". Then turn your attention to the previously-joined timeseries/tsmetrics table and create a subsetted dataframe with only observations of "Atlantic cod" that have a non-NA value. This table reports many years and many metrics, while we are primarily interested in the years since 1950.

-   Which 2 metrics are most commonly reported for Atlantic cod in this timeframe and what do they mean?


```{r}
cod <- stock |> 
  filter(commonname == 'Atlantic cod')

unique(cod$stockid)

unique(cod$region)
  
```

There are 28 unique fish stocks associated with 'Atlantic cod'. They come from the "Canada East Coast", "Europe non EU", "European Union", and "US East Coast" regions.

```{r}
cod_vector <- cod$stockid

time_cod <- time_all |> 
  filter(stockid %in% cod_vector,
         tsyear >= 1950)

time_cod |> 
  group_by(tslong) |> 
  summarize(count = n()) |> 
  arrange(desc(count))
```

The 2 most commonly reported metrics since 1950 are 'catch divided by mean catch' and 'general total Catch (TC then TL, MT units only)'. Since we grouped by the long name of the method, these names are relatively self explainatory. 

### Q6 (1 point)

Make a timeseries figure for each stock of Atlantic cod in a single figure. Rather than coloring by the identity of the stock, color by the region it comes from (this may require joining). Use `facet_*()` to plot the top two metrics from Q5 in a single code chunk. Connect the points for each stock and focus on the period since 1950. Update until the figure is easily interpretable.


-   What happened to Atlantic cod stocks during the period 1950-present? What information does each metric convey?

```{r}
time_cod |> 
  left_join(stock, by = "stockid") |> 
  filter(tslong == "Catch divided by mean catch" | tslong == "General total Catch (TC then TL, MT units only)") |> 
  ggplot(aes(x = tsyear, y = tsvalue, color = region)) +
  geom_point() +
  geom_line(aes(group = stockid)) +
  facet_wrap(~ tslong, scales = "free_y") +
  labs(x = "Year",
       y = "Value",
       color = "Region") +
  theme_bw()
```

Atlantic cod stocks generally decreased during this time period. The first panel shows catch/mean catch, comparing current catch to an average expected catch. The second panel shows general total catch, which is a measure of how many total fish they caught. Both plots are showing decreasing trends in general, except maybe for Europe non EU in general total catch.

### Q7 (1 point)

We will recreate a slightly more complex version of a figure from the [Millennium Ecosystem Assessment](https://www.millenniumassessment.org/documents/document.356.aspx.pdf) (Fig. 11).

Selecting a metric that can be added across stocks, calculate the total fish landings for each region of Atlantic cod. Make a plot similar to Fig. 11 and color by region.


Which region does Fig. 11 from the MEA correspond to?

```{r}
time_cod |> 
  left_join(stock, by = "stockid") |> 
  filter(tslong == "Total landings") |> 
  group_by(region, tsyear) |> 
  summarize(total_landings = sum(tsvalue)) |> 
  ggplot(aes(x = tsyear, y = total_landings, color = region)) +
  geom_point() +
  geom_line() +
  labs(x = "Year",
       y = "Total landings (tons)",
       color = "Region") +
  theme_bw()
```

Figure 11 corresponds to Canada East coast, which is red in our plot. The MEA and our plot (for Canada) both show a spike in total landings between 1960-1970.

### Q8 (1 point)

Under the Management Policies section of the [book chapter](https://pressbooks.pub/extinctionstories/chapter/atlantic-cod/), the authors summarize a few major efforts to restore and protect stocks of Atlantic cod. Focusing only on the US and Canada, update your plot from above to include vertical lines to represent the timing of least three major events and label them clearly on your plot. Doing so will likely involve creating a separate dataframe and using it in one of your ggplot layers (one possible argument within `aes()` is `linetype`). Facet the plot by region and update until it is easily interpretable.

```{r}
important_years <- data.frame(years = c(1976, 1996, 2006), policy = c("MSFCMA",
                                                                      "Sustainable Fisheries Act",
                                                                      "Magnuson-Stevens Fishery Conservation and Management Reauthorization Act"))
time_cod |> 
  left_join(stock, by = "stockid") |> 
  filter(tslong == "Total landings",
         region != "Europe non EU" & region != "European Union") |> 
  group_by(region, tsyear) |> 
  summarize(total_landings = sum(tsvalue)) |> 
  ggplot(aes(x = tsyear, y = total_landings, color = region)) +
  geom_vline(data = important_years, aes(xintercept = years, linetype = policy)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ region, ncol = 1) +
  labs(x = "Year",
       y = "Total landings (tons)",
       color = "Region") +
  theme_bw()
```

### Q9 (2 points)

What happened to Atlantic cod stocks for each region following each piece of legislation or management plan, and why do you think this was so? Your multi-paragraph answer should be informed by content from the book chapter and your figure from Q8; additional sources are optional, and all sources should be cited.

The first piece of legislation we highlighted on our plot was the Magnuson-Stevens Fishery Conservation and Management Act (MSFCMA) in 1976. In our plot, we see that total landings in North America were declining prior to this law, caused by over fishing of deep waters (Brownell et al., 'Extinction Stories'). This law protected US controlled waters from fishing vessels from other countries. We can see a halt in the decline of total landings preceding the legislation.

Despite regulatory efforts, over fishing of Atlantic cod persisted (Brownell et al., 'Extinction Stories'). The species was eventually listed as vulnerable in 1996. Amendments to the MSFCMA were made, notably, the Sustainable Fisheries Act in 1996 and the Magnuson-Stevens Fishery Conservation and Management Reauthorization Act in 2006, both shown on our plot. These Amendments imposed a stricter catch limit and limited fishing in certain habitats vital to the Atlantic cod's proliferation. We do not see a clear response in total cod landings to these pieces of legislation in our data. Total landings was relatively low and remained relatively low before and after these amendments. Studies have shown that despite regulatory efforts, the Atlantic cod populations have not recovered (Brownell et al., 'Extinction Stories'), which is reflected in our data. More efforts for explicit recovery plans rather than just fishing regulations would likely be necessary to revitalize the Atlantic cod populations to their previous abundances. 



